---
title: "MCC RNAseq - variation by origin and viral status"
author: "Keegan Korthauer"
output: 
    html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
        code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
options(bitmapType='cairo')
```

# Summary

Here we take a look at the expression (mRNA-seq) data from the samples sequenced 
thus far for both MCC patient tumors, and their matching patient-derived cell 
lines. 

# Workspace setup

Here we'll load the necessary R packages.

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
theme_set(theme_bw())
library(dplyr)
library(tidyr)
library(DESeq2)
library(readr)
library(readxl)
library(RColorBrewer) 
library(ComplexHeatmap) 
library(vsn)
library(fgsea)
library(biomaRt)
library(BiocParallel)
library(GO.db)
library(org.Hs.eg.db)
library(ggrepel)
library(Gviz)
library(rtracklayer)
library(RUVSeq)
library(cowplot)
library(ggridges)
library(GGally)

register(MulticoreParam(10))

viralquant_file <- "../../PREPROCESS/RNA/viral/virus.read.counts.txt"
```

Now we'll read in the RNA-seq count data, generated with Subread on the bam 
files provided by the Broad's Firecloud (see the `../DATA` directory for the
bam files and download scripts for Firecloud, and see the `../PREPROCESS` 
directory for the count level data and scripts for generating from Rsubread. 
We have saved a single file where each row is a feature and each column is a 
sample:

```{r}
exp <- data.frame(read_tsv("../../PREPROCESS/RNA/mRNA-Seq.counts.txt"))
colnames(exp) <- gsub("X.n.irizarryfs01_backed_up.kkorthauer.MCC.DATA.RNA.", "",
                      colnames(exp))
colnames(exp) <- gsub(".01.bam", "", colnames(exp))
rownames(exp) <- exp[,1]
rowDat <- exp[, c(1:6)]
exp <- exp[, -c(1:6)]
colDat <- data.frame(sampleName = colnames(exp),
                     dfci.id = unlist(lapply(strsplit(colnames(exp), "\\."), 
                                             function(x) paste0(x[1], "_", x[2]))),
                     type = unlist(lapply(strsplit(colnames(exp), "\\."),
                                          function(x) x[3])))
colDat$type <- ifelse(grepl("T", colDat$type), "Tumor", "Cell_line") 
colDat
colDat$type <- as.factor(colDat$type)

table(colDat$dfci.id, colDat$type)
```

Here we have `r sum(rowSums(table(colDat$dfci.id, colDat$type)) == 2)` complete
pairs of tumor and cell line.

We'll also add in the metadata from the de-identified excel file so we can map
the Broad sample IDs back to the original MCC patient ids. Excel file doesn't 
contain one observation per cell, so we'll have to manually enter the information.

```{r}
# 5350 -> 336
# 5351 -> 350
# 5367 -> 277
# 5368 -> 2314
# 5369 -> 301
# 5370 -> 286
# 5474 -> 367
# 5473 -> 320
# 5475 -> 290
# 5476 -> 282
# 5477 -> 275 
# 5478 -> 358

id.map <- rbind(c(5350, 336),
                c(5351, 350),
                c(5367, 277),
                c(5368, 2314),
                c(5369, 301),
                c(5370, 286),
                c(5474, 367),
                c(5473, 320),
                c(5475, 290),
                c(5476, 282),
                c(5477, 275),
                c(5478, 358))
id.map <- data.frame(id.map)
colnames(id.map) <- c("id1", "id2")

x <- match(as.numeric(gsub("DFCI_", "", colDat$dfci.id)), id.map$id1)
colDat$mcc.id <- as.factor(id.map$id2[x])

viralneg <- c("286", "350", "290", "282", "320")
colDat$virus <- "pos"
colDat$virus[colDat$mcc.id %in% viralneg] <- "neg"
colDat$virus <- as.factor(colDat$virus)
```

Finally, we'll construct a DESeq dataset.

```{r}
dat <- DESeqDataSetFromMatrix(countData = exp,
                              colData = colDat,
                              rowData = rowDat,
                              design = ~ type)
```

# Filtering and normalization

Examine lowly expressed / undetected genes.

```{r}
dat <- estimateSizeFactors(dat)
df <- data.frame(Zero = colSums(counts(dat) == 0)/nrow(dat),
                 Low = colSums(counts(dat) < 5)/nrow(dat),
                 Libsize = sizeFactors(dat))
df <- df %>% mutate(sampleName = rownames(df)) %>%
  gather(level, proportion, -sampleName) %>%
  mutate(type = factor(ifelse(grepl("T", sampleName), "Tumor", "Cell Line")))

p1 <- df %>% filter(level != "Libsize") %>%
  ggplot(aes(x = type, y = proportion, fill = type)) +
    facet_grid( ~ level) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.05) +
    theme(legend.position = "none") +
    ggtitle("Proportion of lowly expressed genes")

p2 <- df %>% filter(level == "Libsize") %>%
  ggplot(aes(x = type, y = proportion, fill = type)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.05) +
    ggtitle("Library Size") +
    ylab("Library size")

plot_grid(p1, p2, rel_widths = c(0.7,0.6), nrow = 1)

# t-test for difference between tumor and cell line
t.test(df$proportion[df$level == "Zero"] ~ df$type[df$level ==  "Zero"])
t.test(df$proportion[df$level == "Low"] ~ df$type[df$level ==  "Low"])

# excluding 5370 (286) outlier sample
df <- filter(df, sampleName != "DFCI.5370.T")
t.test(df$proportion[df$level == "Zero"] ~ df$type[df$level ==  "Zero"])
t.test(df$proportion[df$level == "Low"] ~ df$type[df$level ==  "Low"])
```

First we'll remove features (genes) that have mean count across all samples 
less than 1. These are considered not expressed.

```{r}
dat.orig <- dat
filt <- which(rowMeans(counts(dat)) < 1)
dat <- dat[ -filt, ]
```

## Size factor

Next we'll estimate the library size factors.

```{r}
dat <- estimateSizeFactors(dat)
dat.orig <- estimateSizeFactors(dat.orig)
```

Here are boxplots of raw and normalized counts:

```{r}
# raw
data.frame(log(counts(dat)+1)) %>% gather("sampleName", "count", 1:ncol(dat)) %>%
 left_join(data.frame(colData(dat)), by = "sampleName") %>%
 ggplot(aes(x = mcc.id, y = count, group = sampleName, fill = type)) +
 facet_grid( ~ virus, scales = "free_x") +  
 geom_boxplot() +
 ggtitle("Raw counts") +
 ylab("log(raw count + 1)")
ggsave(file.path("plots", "rawcounts_boxplot.pdf"), width = 7, height = 3)

# normalized (size factor) 
data.frame(log(counts(dat, normalize = TRUE)+1)) %>% 
 gather("sampleName", "count", 1:ncol(dat)) %>%
 left_join(data.frame(colData(dat)), by = "sampleName") %>%
 ggplot(aes(x = mcc.id, y = count, group = sampleName, fill = type)) +
 facet_grid( ~ virus, scales = "free_x") +  
 geom_boxplot() +
 ggtitle("Normalized counts") +
 ylab("log(normalized count + 1)")
ggsave(file.path("plots", "normcounts_boxplot.pdf"), width = 7, height = 3)
```

And ridge plots showing the same thing:

```{r}
# raw
data.frame(log(counts(dat)+1)) %>% gather("sampleName", "count", 1:ncol(dat)) %>%
 left_join(data.frame(colData(dat)), by = "sampleName") %>%
 ggplot(aes(y = mcc.id, x = count, group = sampleName, fill = type)) +
 facet_wrap( ~ virus, scales = "free_y") +  
 geom_density_ridges(scale = 0.9, alpha = 0.4)  +
 ggtitle("Raw counts") +
 xlab("log(raw count + 1)")
ggsave(file.path("plots", "rawcounts_ridgeplot.pdf"), width = 7, height = 3)

# normalized
data.frame(log(counts(dat, normalized=TRUE)+1)) %>% gather("sampleName", "count", 1:ncol(dat)) %>%
 left_join(data.frame(colData(dat)), by = "sampleName") %>%
 ggplot(aes(y = mcc.id, x = count, group = sampleName, fill = type)) +
 facet_wrap( ~ virus, scales = "free_y") +  
 geom_density_ridges(scale = 0.9, alpha = 0.4)  +
 ggtitle("Normalized counts") +
 xlab("log(normalized count + 1)")
ggsave(file.path("plots", "normcounts_ridgeplot.pdf"), width = 7, height = 3)
```

## Housekeeping genes

As an alternative to DEseq size factor normalization, we'll also try out normalizing
using a set of housekeeping genes, as the difference in the number of expressed genes
between tumor and cell lines is quite large and could be a confounding factor. First 
we need to map these
gene symbols to ENSEMBL ids. We'll do this using the Bioconductor package 
org.Hs.eg.db

```{r}
map <- select(org.Hs.eg.db,
              keys = unlist(lapply(strsplit(rownames(dat), "\\."), 
				                       function(x) x[1])),
              columns=c("SYMBOL","GENENAME", "ENSEMBL"),
              keytype="ENSEMBL")
x <- match(unlist(lapply(strsplit(rownames(dat), "\\."), function(x) x[1])),
           map$ENSEMBL)
rowData(dat)$symbol <- map$SYMBOL[x]   

# unfiltered
map.orig <- select(org.Hs.eg.db,
              keys = unlist(lapply(strsplit(rownames(dat.orig), "\\."), 
                                       function(x) x[1])),
              columns=c("SYMBOL","GENENAME", "ENSEMBL"),
              keytype="ENSEMBL")
x <- match(unlist(lapply(strsplit(rownames(dat.orig), "\\."), function(x) x[1])),
           map.orig$ENSEMBL)
rowData(dat.orig)$symbol <- map.orig$SYMBOL[x]
```

Next we donwload the list of housekeeping genes.

```{r}
if (!file.exists("../HK_genes.txt")){
  download.file("https://www.tau.ac.il/~elieis/HKG/HK_genes.txt", 
    destfile="../HK_genes.txt")
}

hkgenes <- read.table("../HK_genes.txt",stringsAsFactors = FALSE)
colnames(hkgenes) = c("symbol", "id")

rowData(dat)$housekeeping <- rowData(dat)$symbol %in% hkgenes$symbol

# plot of raw expression of hk genes
data.frame(log(counts(dat)+1)) %>%
 filter(rowData(dat)$housekeeping) %>%
 gather("sampleName", "count", 1:ncol(dat)) %>%
 left_join(data.frame(colData(dat)), by = "sampleName") %>%
 ggplot(aes(x = mcc.id, y = count, group = sampleName, fill = type)) +
 facet_grid( ~ virus, scales = "free_x") +  
 geom_boxplot() +
 ggtitle("Housekeeping genes") +
 ylab("log(raw count + 1)")
```

Use these genes to normalize:

```{r}
set1 <- RUVg(counts(dat), rowData(dat)$housekeeping, k=1)
colData(dat)$W = as.vector(set1$W)

data.frame(log(set1$normalizedCounts+1)) %>%
 gather("sampleName", "count", 1:ncol(dat)) %>%
 left_join(data.frame(colData(dat)), by = "sampleName") %>%
 ggplot(aes(x = mcc.id, y = count, group = sampleName, fill = type)) +
 facet_grid( ~ virus, scales = "free_x") +  
 geom_boxplot() +
 ggtitle("Housekeeping genes") +
 ylab("log(normalized count + 1)")
```

# Unsupervised clustering

Next we'll perform an unsupervised hierarchical clustering of the samples to 
see if there are any similarities among the patients and sample types. The 
heatmaps will be constructed using a distance matrix on variance stabilizing
transformed expression values.

## All genes

We'll first carry out a variance stabilizing transformation, using all genes.

```{r}
# carry out the vs transformation:
vsd = vst(dat)

# plot variance
meanSdPlot(assay(vsd))
```

Pairs scatterplot of all samples

```{r pairscatter, fig.width = 11, fig.height = 10, fig.path='plots/', dev = c('png')}
my_fn <- function(data, mapping, ...){
      p <- ggplot(data = data, mapping = mapping) + 
        #stat_density2d(aes(fill=..density..), geom="tile", contour = FALSE, n = 100) +
        geom_hex(bins = 20) +
        scale_fill_gradientn(colours = viridis::viridis(50)) +
                               #colorRampPalette(c("white", blues9))(50)) +
        geom_abline(intercept=0,slope=1, linetype = "dashed") +
        scale_x_continuous(trans="log10") +
        scale_y_continuous(trans="log10") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1))
      p
}

my_corr <- function (data, mapping, alignPercent = 0.6, method = "pearson", 
    use = "complete.obs", corAlignPercent = NULL, corMethod = NULL, 
    corUse = NULL, ...) 
{
    if (!is.null(corAlignPercent)) {
        stop("'corAlignPercent' is deprecated.  Please use argument 'alignPercent'")
    }
    if (!is.null(corMethod)) {
        stop("'corMethod' is deprecated.  Please use argument 'method'")
    }
    if (!is.null(corUse)) {
        stop("'corUse' is deprecated.  Please use argument 'use'")
    }
    useOptions <- c("all.obs", "complete.obs", "pairwise.complete.obs", 
        "everything", "na.or.complete")
    use <- pmatch(use, useOptions)
    if (is.na(use)) {
        warning("correlation 'use' not found.  Using default value of 'all.obs'")
        use <- useOptions[1]
    }
    else {
        use <- useOptions[use]
    }
    cor_fn <- function(x, y) {
        cor(x, y, method = method, use = use)
    }
    xData <- eval_data_col(data, mapping$x)
    yData <- eval_data_col(data, mapping$y)
    if (GGally:::is_date(xData)) {
        xData <- as.numeric(xData)
    }
    if (GGally:::is_date(yData)) {
        yData <- as.numeric(yData)
    }
    colorData <- eval_data_col(data, mapping$colour)
    if (is.numeric(colorData)) {
        stop("ggally_cor: mapping color column must be categorical, not numeric")
    }
    if (use %in% c("complete.obs", "pairwise.complete.obs", "na.or.complete")) {
        if (!is.null(colorData) && (length(colorData) == length(xData))) {
            rows <- complete.cases(xData, yData, colorData)
        }
        else {
            rows <- complete.cases(xData, yData)
        }
        if (any(!rows)) {
            total <- sum(!rows)
            if (total > 1) {
                warning("Removed ", total, " rows containing missing values")
            }
            else if (total == 1) {
                warning("Removing 1 row that contained a missing value")
            }
        }
        if (!is.null(colorData) && (length(colorData) == length(xData))) {
            colorData <- colorData[rows]
        }
        xData <- xData[rows]
        yData <- yData[rows]
    }
    xVal <- xData
    yVal <- yData
    if (packageVersion("ggplot2") > "2.2.1") {
        for (mappingName in names(mapping)) {
            itemData <- eval_data_col(data, mapping[[mappingName]])
            if (!inherits(itemData, "AsIs")) {
                mapping[[mappingName]] <- NULL
            }
        }
    }
    else {
        if (length(names(mapping)) > 0) {
            for (i in length(names(mapping)):1) {
                tmp_map_val <- deparse(mapping[names(mapping)[i]][[1]])
                if (tmp_map_val[length(tmp_map_val)] %in% colnames(data)) 
                  mapping[[names(mapping)[i]]] <- NULL
                if (length(names(mapping)) < 1) {
                  mapping <- NULL
                  break
                }
            }
        }
    }
    if (!is.null(colorData) && !inherits(colorData, "AsIs")) {
        cord <- ddply(data.frame(x = xData, y = yData, color = colorData), 
            "color", function(dt) {
                cor_fn(dt$x, dt$y)
            })
        colnames(cord)[2] <- "correlation"
        cord$correlation <- signif(as.numeric(cord$correlation), 
            3)
        lev <- levels(as.factor(colorData))
        ord <- rep(-1, nrow(cord))
        for (i in 1:nrow(cord)) {
            for (j in seq_along(lev)) {
                if (identical(as.character(cord$color[i]), as.character(lev[j]))) {
                  ord[i] <- j
                }
            }
        }
        cord <- cord[order(ord[ord >= 0]), ]
        cord$label <- str_c(cord$color, ": ", cord$correlation)
        xmin <- min(xVal, na.rm = TRUE)
        xmax <- max(xVal, na.rm = TRUE)
        xrange <- c(xmin - 0.01 * (xmax - xmin), xmax + 0.01 * 
            (xmax - xmin))
        ymin <- min(yVal, na.rm = TRUE)
        ymax <- max(yVal, na.rm = TRUE)
        yrange <- c(ymin - 0.01 * (ymax - ymin), ymax + 0.01 * 
            (ymax - ymin))
        p <- ggally_text(label = str_c(signif(cor_fn(xVal, 
            yVal), 2)), mapping = mapping, xP = 0.5, yP = 0.9, 
            xrange = xrange, yrange = yrange, color = "black", 
            ...) + theme(legend.position = "none", 
                         panel.grid.major = element_blank()) 
        xPos <- rep(alignPercent, nrow(cord)) * diff(xrange) + 
            min(xrange, na.rm = TRUE)
        yPos <- seq(from = 0.9, to = 0.2, length.out = nrow(cord) + 
            1)
        yPos <- yPos * diff(yrange) + min(yrange, na.rm = TRUE)
        yPos <- yPos[-1]
        cordf <- data.frame(xPos = xPos, yPos = yPos, labelp = cord$label)
        cordf$labelp <- factor(cordf$labelp, levels = cordf$labelp)
        p <- p + geom_text(data = cordf, aes(x = xPos, y = yPos, 
            label = labelp, color = labelp), hjust = 1, ...) +
            theme(panel.grid.major = element_blank())
        p
    }
    else {
        xmin <- min(xVal, na.rm = TRUE)
        xmax <- max(xVal, na.rm = TRUE)
        xrange <- c(xmin - 0.01 * (xmax - xmin), xmax + 0.01 * 
            (xmax - xmin))
        ymin <- min(yVal, na.rm = TRUE)
        ymax <- max(yVal, na.rm = TRUE)
        yrange <- c(ymin - 0.01 * (ymax - ymin), ymax + 0.01 * 
            (ymax - ymin))
        p <- ggally_text(label = paste(signif(cor_fn(xVal, 
            yVal), 2), sep = "", collapse = ""), mapping, xP = 0.5, 
            yP = 0.5, xrange = xrange, yrange = yrange, ...) + 
            theme(legend.position = "none", panel.grid.major = element_blank())
        p
    }
}

plotpairs <- function(mat, main = ""){
  p1 <- ggpairs(mat,
                lower = list(continuous = my_fn),
    upper = list(continuous = wrap(my_corr, size = 6, 
                                   color = "black", method = "spearman")),
    xlab = "Normalized Read Count", 
    ylab = "Normalized Read Count",
    title = main) + 
    theme(axis.text.x = element_text(angle=90, hjust = 1))
  p2 <- ggcorr(mat, label = TRUE, label_round = 2, limits = c(0.5, 1), 
               midpoint = 0.75, name ="Spearman\nCorrelation", 
               method = c("pairwise", "spearman"))
  leg <- get_legend(p2) 
  
  g2 <- ggplotGrob(p2)
  colors <- g2$grobs[[6]]$children[[3]]$gp$fill
  
  # Change background color to tiles in the upper triangular matrix of plots 
  idx <- 1
  p <- ncol(mat)
  for (k1 in 1:(p-1)) {
    for (k2 in (k1+1):p) {
      plt <- getPlot(p1,k1,k2) +
        theme(panel.background = element_rect(fill = colors[idx], color="white"),
              panel.grid.major = element_line(color=colors[idx]))
      p1 <- putPlot(p1,plt,k1,k2)
      idx <- idx+1
    }
  }
  
  for (i in 1:p){ # log scale for the diagonal
      p1[i,i] <- p1[i,i]+ scale_x_continuous(trans='log10')  
  } 

  
  p1 <- plot_grid(ggmatrix_gtable(p1), leg, ncol=2, rel_widths = c(1, 0.2))
  print(p1)
}


mat <- data.frame(counts(dat, normalize = TRUE)+1)
x2 <- match(as.numeric(gsub(".C|.T|.CL", "", gsub("DFCI.", "", colnames(mat)))), id.map$id1)
colnames(mat) <- paste0(ifelse(grepl("T", colnames(mat)), "T-", "C-"), id.map$id2[x2])


# just for tumors
plotpairs(mat[, grepl("T", colnames(mat))], 
          main = "Correlation of Tumor Samples")

# just for cell lines
plotpairs(mat[, !grepl("T", colnames(mat))], 
          main = "Correlation of Cell Line Samples")

# just for virus pos
plotpairs(mat[, !grepl(paste0(viralneg, collapse = "|"), colnames(mat))], 
          main = "Correlation of Virus+ Samples")

# just for virus neg
plotpairs(mat[, grepl(paste0(viralneg, collapse = "|"), colnames(mat))], 
          main = "Correlation of Virus- Samples")

# for everything
plotpairs(mat, 
          main = "Correlation of All Samples")

```

Now we'll plot the distances in a heatmap, first for log2 normalized counts:


```{r heatmap-dist-l2n, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
dists <- dist(t(log2(counts(dat, normalize = TRUE)+1)))
dt <- function(x){
 return(as.dist(x))
}
mat <- as.matrix(dists) # convert to matrix
rownames(mat) <- colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", 
                                       gsub("_", " ", type))) 

colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255) # set colors
cols2 <-  colorRampPalette( rev(brewer.pal(length(unique(colData(dat)$mcc.id)), 
                                           "Set3")) )(length(unique(colData(dat)$mcc.id)))
names(cols2) <- unique(colData(dat)$mcc.id)
ha_column = HeatmapAnnotation(df = data.frame(Type = colData(dat)[, c("type")],
                                              Virus = colData(dat)[, c("virus")],
                                              ID = colData(dat)[, c("mcc.id")]),
                              col = list(Type = c(Cell_line="lightgrey", Tumor="black"), 
                                         Virus = c(pos = "#543005", neg = "#DFC27D"),
                                         ID = cols2))
ht = Heatmap(mat, name = "Distance\n(log2 Norm)", 
             top_annotation = ha_column, col = colors,
             clustering_distance_rows=dt,
             clustering_distance_columns=dt)
draw(ht)

```


Next for variance-stabilizing transformed counts:

```{r heatmap-dist-vs, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
distsVS <- dist(t(assay(vsd)))
dt <- function(x){
 return(as.dist(x))
}
mat <- as.matrix(distsVS) # convert to matrix
rownames(mat) <- colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", 
                                       gsub("_", " ", type))) 
ha_right = rowAnnotation(df = data.frame(ID = colData(dat)[, c("mcc.id")]),
                             col = list(ID = cols2),
                             show_legend=FALSE)
ht = Heatmap(mat, name = "Distance\n(VS)", 
             top_annotation = ha_column, col = colors,
             #right_annotation = ha_right,
             clustering_distance_rows=dt,
             clustering_distance_columns=dt,
             show_row_names = FALSE,
             show_column_names = FALSE)
draw(ht + ha_right)
```

Same as previous, but remove outlier sample

```{r heatmap-dist-vs-no286, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
distsVS <- dist(t(assay(vsd)[,!grepl("5370", colnames(assay(vsd)))]))
dt <- function(x){
 return(as.dist(x))
}
mat <- as.matrix(distsVS) # convert to matrix
rownames(mat) <- colnames(mat) <- with(colData(dat)[colData(dat)$mcc.id != "286",], paste0(mcc.id, " ", 
                                       gsub("_", " ", type))) 
ha_right = rowAnnotation(df = data.frame(ID = colData(dat)[colData(dat)$mcc.id != "286", c("mcc.id")]),
                             col = list(ID = cols2),
                             show_legend=FALSE)
ha_column = HeatmapAnnotation(df = data.frame(Type = colData(dat)[colData(dat)$mcc.id != "286", c("type")],
                                              Virus = colData(dat)[colData(dat)$mcc.id != "286", c("virus")],
                                              ID = colData(dat)[colData(dat)$mcc.id != "286", c("mcc.id")]),
                              col = list(Type = c(Cell_line="lightgrey", Tumor="black"), 
                                         Virus = c(pos = "#543005", neg = "#DFC27D"),
                                         ID = cols2))
ht = Heatmap(mat, name = "Distance\n(VS)", 
             top_annotation = ha_column, col = colors,
             #right_annotation = ha_right,
             clustering_distance_rows=dt,
             clustering_distance_columns=dt,
             show_row_names = FALSE,
             show_column_names = FALSE)
draw(ht+ha_right)
```


## Highly variable genes 

Next we'll use only a smaller set of variable genes. First for log2 normalized counts:

```{r heatmap-dist-var-l2n, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
ha_column = HeatmapAnnotation(df = data.frame(Type = colData(dat)[, c("type")],
                                              Virus = colData(dat)[, c("virus")],
                                              ID = colData(dat)[, c("mcc.id")]),
                              col = list(Type = c(Cell_line="lightgrey", Tumor="black"), 
                                         Virus = c(pos = "#543005", neg = "#DFC27D"),
                                         ID = cols2))


# rank variable genes:
top <- rank(-rowVars(counts(dat)), ties.method = "first")

# Calculate distances using transformed (and normalized) counts
# only for top 1000 genes
dists <- dist(t(log2(counts(dat, normalize = TRUE)[top <= 1000,]+1)))
mat <- as.matrix(dists) # convert to matrix
rownames(mat) <- colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", 
                                       gsub("_", " ", type))) 

ht = Heatmap(mat, name = "Distance\n(log2 Norm)", 
             top_annotation = ha_column, col = colors,
             clustering_distance_rows=dt,
             clustering_distance_columns=dt)
draw(ht)
```     

Next for variance-stabilizing transformed counts:

```{r heatmap-dist-var-vs, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
# rank variable genes:
top <- rank(-rowVars(assay(vsd)), ties.method = "first")

# Calculate distances using transformed (and normalized) counts
# only for top 1000 genes
distsVS <- dist(t(assay(vsd)[top <= 1000,]))
mat <- as.matrix(distsVS) # convert to matrix
rownames(mat) <- colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", 
                                       gsub("_", " ", type))) 

ht = Heatmap(mat, name = "Distance\n(VS)", 
             top_annotation = ha_column, col = colors,
             clustering_distance_rows=dt,
             clustering_distance_columns=dt)
draw(ht)
```     

## Visualize highly variable genes

Here we'll create a heatmap of genes (rows) by samples (columns) of the most
highly variable genes (instead of the distance as done in the previous 
subsections). First for log2 normalized counts:

```{r heatmap-expr-var-100-l2n, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
# Calculate distances using transformed (and normalized) counts
# only for top 100 genes
mat <- log2(counts(dat, normalize = TRUE)[top <= 100,]+1)
colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", gsub("_", " ", type))) 

ecolors <- c("white", colorRampPalette( (brewer.pal(9, "Reds")) )(255)) # set colors
ht = Heatmap(mat, name = "log2\nNormalized\nCounts", 
             top_annotation = ha_column,
             show_row_names = FALSE,
             clustering_distance_rows = "euclidean",
             clustering_distance_columns = "euclidean",
             col = ecolors)
draw(ht)
```     

Next for variance-stabilizing transformed counts:

```{r heatmap-expr-var-100-vs, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
# Calculate distances using transformed (and normalized) counts
# only for top 100 genes
mat <- assay(vsd)[top <= 100,]
colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS Counts", 
             top_annotation = ha_column,
             show_row_names = FALSE,
             clustering_distance_rows = "euclidean",
             clustering_distance_columns = "euclidean",
             col = ecolors)
draw(ht)
```     

Instead of 100 highly variable genes, we also show results for 500. This time,
also exclude the virus negative samples.

```{r heatmap-expr-var-500-vs-vp, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
dat_p <- dat[, c(which(colData(dat)$virus == "pos"))]
vsd_p = vst(dat_p)

cols2_p <-  colorRampPalette( rev(brewer.pal(length(unique(colData(dat_p)$mcc.id)), 
                                           "Set3")) )(length(unique(colData(dat_p)$mcc.id)))
names(cols2_p) <- unique(colData(dat_p)$mcc.id)
ha_column_p = HeatmapAnnotation(df = data.frame(Type = colData(dat_p)[, c("type")],
                                                ID = colData(dat_p)[, c("mcc.id")]),
                               col = list(Type = c(Cell_line="lightgrey", Tumor="black"), 
                                          ID = cols2_p))

# Calculate distances using transformed (and normalized) counts
# only for top 500 genes
top <- rank(-rowVars(assay(vsd_p)), ties.method = "first")
mat <- assay(vsd_p)[top <= 500,]
colnames(mat) <- with(colData(dat_p), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column_p,
             show_row_names = FALSE,
             col = ecolors)
draw(ht)
```     

And now the same for viral negative samples:
        
```{r heatmap-expr-var-500-vp, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
dat_n <- dat[, c(which(colData(dat)$virus %in% "neg"))]
vsd_n = vst(dat_n)

cols2_n <-  colorRampPalette( rev(brewer.pal(length(unique(colData(dat_n)$mcc.id)), 
                                           "Set3")) )(length(unique(colData(dat_n)$mcc.id)))
names(cols2_n) <- unique(colData(dat_n)$mcc.id)
ha_column_n = HeatmapAnnotation(df = data.frame(Type = colData(dat_n)[, c("type")],
                                                ID = colData(dat_n)[, c("mcc.id")]),
                               col = list(Type = c(Cell_line="lightgrey", Tumor="black"), 
                                          ID = cols2_n))

# Calculate distances using transformed (and normalized) counts
# only for top 500 genes
top <- rank(-rowVars(assay(vsd_n)), ties.method = "first")
mat <- assay(vsd_n)[top <= 500,]
colnames(mat) <- with(colData(dat_n), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column_n,
             show_row_names = FALSE,
             col = ecolors)
draw(ht)
```     
        
        
# Principal Components 

Here we'll make a principal components plot as another view of how similar 
the different samples are to one another.

```{r}
pcaData <- plotPCA(vsd, intgroup=c("type", "mcc.id"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=type, shape=mcc.id)) +
  scale_shape_manual(values=1:nlevels(pcaData$mcc.id)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
ggsave(file.path("plots", "pca_vscounts.pdf"), width = 5, height = 4)
```        

Now colored by viral status:

```{r}
pcaData <- plotPCA(vsd, intgroup=c("type", "mcc.id", "virus"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=virus, shape=mcc.id)) +
  scale_shape_manual(values=1:nlevels(pcaData$mcc.id)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
ggsave(file.path("plots", "pca_vscounts_viral.pdf"), width = 5, height = 4)
```        

And the same PCA plot constructed on the top 500 genes, for only the subset of
samples that are virus positive:

```{r}
pcaData <- plotPCA(vsd_p, intgroup=c("type", "mcc.id"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=type, shape=mcc.id)) +
  scale_shape_manual(values=1:nlevels(pcaData$mcc.id)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
ggsave(file.path("plots", "pca_vscounts_subset.pdf"), width = 5, height = 4)
```              

Next for the virus-negative samples:

```{r}
pcaData <- plotPCA(vsd_n, intgroup=c("type", "mcc.id"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=type, shape=mcc.id)) +
  scale_shape_manual(values=1:nlevels(pcaData$mcc.id)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
ggsave(file.path("plots", "pca_vscounts_subset.pdf"), width = 5, height = 4)
```              
         
# Characterizing immune genes of interest

We'll investigate specific genes of interest, including HLA genes. The list of 
genes is provided in a tab-delimited text file where the first column contains 
gene symbols and the second contains a description.

```{r}
goi <- read_tsv("../genes_of_interest.txt", col_names = FALSE)
colnames(goi) <- c("symbol", "description")
```

```{r}
# add description to rowdata
x <- which(rowData(dat)$symbol %in% goi$symbol)

rowData(dat)$description <- NA
for(g in x){
  rowData(dat)$description[g] <- goi$description[goi$symbol == rowData(dat)$symbol[g]]
}
```

Next, we'll create a heatmap of these genes of interest, first for log2 normalized
counts:


```{r heatmap-goi-l2n, fig.width=8.5, fig.height=10, fig.path='plots/', dev=c('png')}
# indices of genes of interest and extra HLA genes
x <- which(rowData(dat)$symbol %in% goi$symbol)      

# subset normalized counts
mat <- log2(counts(dat, normalize = TRUE)[x,] + 1)
rownames(mat) <- rowData(dat)$symbol[x]
colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", gsub("_", " ", type))) 

rowcol <- colorRampPalette( rev(brewer.pal(10, "Paired")) )(10)
#rowcol <- rev(c("#8dd3c7", "#bebada", "#ffffb3", "#fb8072"))
names(rowcol) <- sort(unique(rowData(dat)$description[!is.na(rowData(dat)$description)]))
ha_row = rowAnnotation(df = data.frame(Category = rowData(dat)[x,"description"]),
                       col = list(Category = rowcol))

ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column,
             row_names_gp = gpar(fontsize = 7),
             col = ecolors)
draw(ht + ha_row)
```

Next for vs-transformed counts.

```{r heatmap-goi-vs, fig.width=8.5, fig.height=10, fig.path='plots/', dev=c('png')}
# subset counts of transformed (and normalized) counts
mat <- assay(vsd)[x,]
rownames(mat) <- rowData(dat)$symbol[x]
colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column,
             row_names_gp = gpar(fontsize = 7),
             col = ecolors)
draw(ht + ha_row)
```

## Subset of genes for Lucas

We'll investigate only Class I and Class II genes, with the addition of Polycomb and housekeeping genes provided by Lucas.

```{r}
goi <- read_tsv("../genes_of_interest.txt", col_names = FALSE)
goi <- goi[,c(1:2)]
colnames(goi) <- c("symbol", "description")

# filter on just class I/II
goi <- goi[goi$description %in% c("Class I", "Class II"),]
```

Add extra PcG and Housekeeping genes. Per Lucas' email 7/30/2019:
>PcG: PCGF1, BCORL1, USP7, BCOR, ASXL1, BAP1, CXorf67
>Housekeeping: ACTB, GAPDH

```{r}
goi <- rbind(goi,
             c("PCGF1", "PcG"),
             c("BCORL1", "PcG"),
             c("USP7", "PcG"),
             c("BCOR", "PcG"),
             c("ASXL1", "PcG"),
             c("BAP1", "PcG"),
             c("C10orf67", "PcG"),
             c("ACTB", "Housekeeping"),
             c("GAPDH", "Housekeeping"))
```

```{r}
# add description to rowdata
x <- which(rowData(dat.orig)$symbol %in% goi$symbol)

rowData(dat.orig)$description <- NA
for(g in x){
  rowData(dat.orig)$description[g] <- goi$description[goi$symbol == rowData(dat.orig)$symbol[g]]
}
```

Next, we'll create a heatmap of these genes of interest, first for log2 normalized
counts:


```{r heatmap-goi-l2n-lucas, fig.width=8.5, fig.height=10, fig.path='plots/', dev=c('png')}
# indices of genes of interest and extra HLA genes
x <- which(rowData(dat.orig)$symbol %in% goi$symbol)      

# subset normalized counts
mat <- log2(counts(dat.orig, normalize = TRUE)[x,] + 1)
rownames(mat) <- rowData(dat.orig)$symbol[x]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

rowcol <- colorRampPalette( rev(brewer.pal(4, "Paired")) )(4)
#rowcol <- rev(c("#8dd3c7", "#bebada", "#ffffb3", "#fb8072"))
names(rowcol) <- sort(unique(rowData(dat.orig)$description[!is.na(rowData(dat.orig)$description)]))
ha_row = rowAnnotation(df = data.frame(Category = rowData(dat.orig)[x,"description"]),
                       col = list(Category = rowcol))

ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column,
             row_names_gp = gpar(fontsize = 7),
             col = ecolors)
draw(ht + ha_row)
```

Next for vs-transformed counts.

```{r heatmap-goi-vs-lucas, fig.width=8.5, fig.height=10, fig.path='plots/', dev=c('png')}
# subset counts of transformed (and normalized) counts
mat <- assay(vst(dat.orig))[x,]
rownames(mat) <- rowData(dat.orig)$symbol[x]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column,
             row_names_gp = gpar(fontsize = 7),
             col = ecolors)
draw(ht + ha_row)
```

## Subset of genes for Patrick

We'll repeat with the pared-down list of immune genes sent by Patrick via the Dropbox folder ("RNA-Seq and WGBS Gene Set for Class I APM.xlsx").


```{r}
goi <- read_xlsx("../RNA-Seq and WGBS Gene Set for Class I APM.xlsx", col_names=FALSE)
colnames(goi) <- c("symbol")

symbol <- goi$symbol
```

The following gene symbols from this list are not found in the transcriptome set: `r 
symbol[which(!(symbol %in% rowData(dat.orig)$symbol ))]`. 

After a quick search, we'll add in *PDIA2* and *CANX* to represent these symbols. 

```{r heatmap-goi-l2n-patrick, fig.width=8.5, fig.height=10, fig.path='plots/', dev=c("png")}

symbol <- c(goi$symbol, "PDIA2", "CANX")

# indices of genes of interest and extra HLA genes
x <- which(rowData(dat.orig)$symbol %in% symbol)      

# subset normalized counts
mat <- log2(counts(dat.orig, normalize = TRUE)[x,] + 1)
rownames(mat) <- rowData(dat.orig)$symbol[x]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column,
             row_names_gp = gpar(fontsize = 7),
             col = ecolors)
draw(ht)
```

Next for vs-transformed counts.

```{r heatmap-goi-vs-patrick, fig.width=8.5, fig.height=10, fig.path='plots/', dev=c("png")}
# subset counts of transformed (and normalized) counts
mat <- assay(vst(dat.orig))[x,]
rownames(mat) <- rowData(dat.orig)$symbol[x]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column,
             row_names_gp = gpar(fontsize = 7),
             col = ecolors)
draw(ht)
```


### Revised subset of Class I genes for Patrick

We'll repeat with the pared-down list of Class I genes sent by Patrick via the Dropbox folder ("UPDATED 12-23-19 RNA-Seq and WGBS Gene Set for Class I APM (1).xlsx").

```{r}
goi <- read_xlsx("../UPDATED 12-23-19 RNA-Seq and WGBS Gene Set for Class I APM (1).xlsx", col_names=FALSE)
colnames(goi) <- c("symbol")

symbol <- goi$symbol
```

```{r heatmap-goi-l2n-patrick-revised, fig.width=6, fig.height=8, fig.path='plots/', dev=c("png")}
# indices of genes of interest and extra HLA genes
x <- which(rowData(dat.orig)$symbol %in% symbol)  

# subset normalized counts
mat <- log2(counts(dat.orig, normalize = TRUE)[x,] + 1)
rownames(mat) <- rowData(dat.orig)$symbol[x]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column,
             row_names_gp = gpar(fontsize = 7),
             col = ecolors, column_dend_reorder=TRUE,
             row_dend_reorder=TRUE)

draw(ht)
```

Next for vs-transformed counts.

```{r heatmap-goi-vs-patrick-revised, fig.width=6, fig.height=8, fig.path='plots/', dev=c("png")}
# subset counts of transformed (and normalized) counts
mat <- assay(vst(dat.orig))[x,]
rownames(mat) <- rowData(dat.orig)$symbol[x]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column,
             row_names_gp = gpar(fontsize = 7),
             col = ecolors,column_dend_reorder=TRUE,
             row_dend_reorder=TRUE)

draw(ht)
```


# Characterizing ORF screen genes of interest

First we read in the spreadsheet from Patrick with the top 100 positive and negative gene hits for the ORF screen.

```{r}
orf <- read_xlsx("../MCC-301 ORF Screen Gene List.xlsx")
```


```{r}
# add description to rowdata
x1 <- which(rowData(dat)$symbol %in% orf$'Negative Hits')
x2 <- which(rowData(dat)$symbol %in% orf$'Positive Hits')

rowData(dat)$orf <- "None"
rowData(dat)$orf[x1] <- "Negative"
rowData(dat)$orf[x2] <- "Positive"

rowData(dat)$negative <- NA
for(g in x1){
  rowData(dat)$negative[g] <- orf$'Negative Hits'[orf$'Negative Hits' == rowData(dat)$symbol[g]]
}

rowData(dat)$positive <- NA
for(g in x2){
  rowData(dat)$positive[g] <- orf$'Positive Hits'[orf$'Positive Hits' == rowData(dat)$symbol[g]]
}

df <- data.frame(log(counts(dat, normalized=TRUE)+1)) %>% 
 mutate(orf = rowData(dat)$orf) %>%
 gather("sampleName", "count", 1:ncol(dat)) %>%
 left_join(data.frame(colData(dat)), by = "sampleName")

df %>%
 ggplot(aes(x = mcc.id, y = count, fill = orf))  +
  geom_boxplot() +
 facet_grid(type ~ .) +  
 ggtitle("Normalized counts") +
 ylab("log(normalized count + 1)") +
 xlab("MCC id")
ggsave(file.path("plots", "normcounts_boxplot_byORFstatus.pdf"), width = 9, height = 4)

#### for all genes no filtering
# add gene symbol to dat.orig
map <- select(org.Hs.eg.db,
              keys = unlist(lapply(strsplit(rownames(dat.orig), "\\."), 
				                       function(x) x[1])),
              columns=c("SYMBOL","GENENAME", "ENSEMBL"),
              keytype="ENSEMBL")
x <- match(unlist(lapply(strsplit(rownames(dat.orig), "\\."), function(x) x[1])),
           map$ENSEMBL)
rowData(dat.orig)$symbol <- map$SYMBOL[x]   

# add description to rowdata
x1 <- which(rowData(dat.orig)$symbol %in% orf$'Negative Hits')
x2 <- which(rowData(dat.orig)$symbol %in% orf$'Positive Hits')

rowData(dat.orig)$orf <- "None"
rowData(dat.orig)$orf[x1] <- "Negative"
rowData(dat.orig)$orf[x2] <- "Positive"

rowData(dat.orig)$negative <- NA
for(g in x1){
  rowData(dat.orig)$negative[g] <- orf$'Negative Hits'[orf$'Negative Hits' == rowData(dat.orig)$symbol[g]]
}

rowData(dat.orig)$positive <- NA
for(g in x2){
  rowData(dat.orig)$positive[g] <- orf$'Positive Hits'[orf$'Positive Hits' == rowData(dat.orig)$symbol[g]]
}

df <- data.frame(log(counts(dat.orig, normalized=TRUE)+1)) %>% 
 mutate(orf = rowData(dat.orig)$orf) %>%
 gather("sampleName", "count", 1:ncol(dat.orig)) %>%
 left_join(data.frame(colData(dat.orig)), by = "sampleName")

df %>%
 ggplot(aes(x = mcc.id, y = count, fill = orf))  +
  geom_boxplot() +
 facet_grid(type ~ .) +  
 ggtitle("Normalized counts") +
 ylab("log(normalized count + 1)") +
 xlab("MCC id")
ggsave(file.path("plots", "normcounts_boxplot_byORFstatus_allgenes.pdf"), width = 9, height = 4)
```

Next, we'll create a heatmap of these genes of interest, first for log2 normalized
counts:


```{r heatmap-orf-l2n, fig.width=8.5, fig.height=10, fig.path='plots/', dev=c('png')}
x <- which(rowData(dat.orig)$orf %in% c("Negative", "Positive"))
rN <- match(orf$`Negative Hits`, rowData(dat.orig)$symbol[x])
rN <- rN[!is.na(rN)]
rP <- match(orf$`Positive Hits`, rowData(dat.orig)$symbol[x])
rP <- rP[!is.na(rP)]

# reorder cols
dat.orig <- dat.orig[,  with(colData(dat.orig), order(type, virus, mcc.id))]

# subset normalized counts
mat <- log2(counts(dat.orig, normalize = TRUE)[x,][c(rP, rN),] + 1)
rownames(mat) <- rowData(dat.orig)$symbol[x][c(rP, rN)]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

rowcol <- c("#8dd3c7", "#fb8072")
names(rowcol) <- sort(unique(rowData(dat.orig)$orf[x][c(rP, rN)]))
ha_row = rowAnnotation(df = data.frame(Category = rowData(dat.orig)[x,"orf"][c(rP,rN)]),
                       col = list(Category = rowcol))

ha_column_orf = HeatmapAnnotation(df = data.frame(Type = colData(dat.orig)[, c("type")],
                                              Virus = colData(dat.orig)[, c("virus")],
                                              ID = colData(dat.orig)[, c("mcc.id")]),
                              col = list(Type = c(Cell_line="lightgrey", Tumor="black"), 
                                         Virus = c(pos = "#543005", neg = "#DFC27D"),
                                         ID = cols2))

# cluster rows/cols
ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column_orf,
             row_names_gp = gpar(fontsize = 4),
             col = ecolors)
draw(ht + ha_row)

# don't cluster rows/cols
ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column_orf,
             row_names_gp = gpar(fontsize = 4),
             col = ecolors,
             cluster_columns = FALSE,
             column_order = seq_len(ncol(dat.orig)),
             cluster_rows = FALSE)
draw(ht + ha_row)
```

Next for vs-transformed counts.

```{r heatmap-orf-vs, fig.width=9.5, fig.height=14, fig.path='plots/', dev=c('png')}
# subset counts of transformed (and normalized) counts
mat <- assay(vst(dat.orig))[x,][c(rP, rN),]
rownames(mat) <- rowData(dat.orig)$symbol[x][c(rP, rN)]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column_orf,
             row_names_gp = gpar(fontsize = 3),
             col = ecolors)
draw(ht + ha_row)

# don't cluster rows/cols
ht = Heatmap(mat, name = "VS", 
             top_annotation = ha_column_orf,
             row_names_gp = gpar(fontsize = 4),
             col = ecolors,
             cluster_columns = FALSE,
             column_order = seq_len(ncol(dat.orig)),
             cluster_rows = FALSE)
draw(ht + ha_row)
```


# Characterizing CRISPR screen genes of interest

First we read in the spreadsheet from Patrick with the top and bottom 100 gene hits for the CRISPR screen.

```{r}
crispr <- read_xlsx("../MHC CRISPR Screen Top 100 Hits.xlsx", skip = 1, col_names = TRUE)
```


```{r}
# add description to rowdata
x1 <- which(rowData(dat)$symbol %in% crispr$'Top Hits')
x2 <- which(rowData(dat)$symbol %in% crispr$'Bottom Hits')

rowData(dat)$crispr <- "None"
rowData(dat)$crispr[x1] <- "Top"
rowData(dat)$crispr[x2] <- "Bottom"

rowData(dat)$Top <- NA
for(g in x1){
  rowData(dat)$Top[g] <- crispr$'Top Hits'[crispr$'Top Hits' == rowData(dat)$symbol[g]]
}

rowData(dat)$Bottom <- NA
for(g in x2){
  rowData(dat)$Bottom[g] <- crispr$'Bottom Hits'[crispr$'Bottom Hits' == rowData(dat)$symbol[g]]
}

df <- data.frame(log(counts(dat, normalized=TRUE)+1)) %>% 
 mutate(crispr = rowData(dat)$crispr) %>%
 gather("sampleName", "count", 1:ncol(dat)) %>%
 left_join(data.frame(colData(dat)), by = "sampleName")

df %>%
 ggplot(aes(x = mcc.id, y = count, fill = crispr))  +
  geom_boxplot() +
 facet_grid(type ~ .) +  
 ggtitle("Normalized counts") +
 ylab("log(normalized count + 1)") +
 xlab("MCC id")
ggsave(file.path("plots", "normcounts_boxplot_byCRISPRstatus.pdf"), width = 9, height = 4)

#### for all genes no filtering
# add gene symbol to dat.orig
map <- select(org.Hs.eg.db,
              keys = unlist(lapply(strsplit(rownames(dat.orig), "\\."), 
				                       function(x) x[1])),
              columns=c("SYMBOL","GENENAME", "ENSEMBL"),
              keytype="ENSEMBL")
x <- match(unlist(lapply(strsplit(rownames(dat.orig), "\\."), function(x) x[1])),
           map$ENSEMBL)
rowData(dat.orig)$symbol <- map$SYMBOL[x]   

# add description to rowdata
x1 <- which(rowData(dat.orig)$symbol %in% crispr$'Top Hits')
x2 <- which(rowData(dat.orig)$symbol %in% crispr$'Bottom Hits')

rowData(dat.orig)$crispr <- "None"
rowData(dat.orig)$crispr[x1] <- "Top"
rowData(dat.orig)$crispr[x2] <- "Bottom"

rowData(dat.orig)$Top <- NA
for(g in x1){
  rowData(dat.orig)$Top[g] <- crispr$'Top Hits'[crispr$'Top Hits' == rowData(dat.orig)$symbol[g]]
}

rowData(dat.orig)$Bottom <- NA
for(g in x2){
  rowData(dat.orig)$Bottom[g] <- crispr$'Bottom Hits'[crispr$'Bottom Hits' == rowData(dat.orig)$symbol[g]]
}

df <- data.frame(log(counts(dat.orig, normalized=TRUE)+1)) %>% 
 mutate(crispr = rowData(dat.orig)$crispr) %>%
 gather("sampleName", "count", 1:ncol(dat.orig)) %>%
 left_join(data.frame(colData(dat.orig)), by = "sampleName")

df %>%
 ggplot(aes(x = mcc.id, y = count, fill = crispr))  +
  geom_boxplot() +
 facet_grid(type ~ .) +  
 ggtitle("Normalized counts") +
 ylab("log(normalized count + 1)") +
 xlab("MCC id")
ggsave(file.path("plots", "normcounts_boxplot_byCRISPRstatus_allgenes.pdf"), width = 9, height = 4)
```

Next, we'll create a heatmap of these genes of interest, first for log2 normalized
counts:


```{r heatmap-crispr-l2n, fig.width=8.5, fig.height=10, fig.path='plots/', dev=c('png')}
x <- which(rowData(dat.orig)$crispr %in% c("Top", "Bottom"))
rN <- match(crispr$`Top Hits`, rowData(dat.orig)$symbol[x])
rN <- rN[!is.na(rN)]
rP <- match(crispr$`Bottom Hits`, rowData(dat.orig)$symbol[x])
rP <- rP[!is.na(rP)]

# reorder cols
dat.orig <- dat.orig[,  with(colData(dat.orig), order(type, virus, mcc.id))]

# subset normalized counts
mat <- log2(counts(dat.orig, normalize = TRUE)[x,][c(rP, rN),] + 1)
rownames(mat) <- rowData(dat.orig)$symbol[x][c(rP, rN)]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

rowcol <- c("#8dd3c7", "#fb8072")
names(rowcol) <- sort(unique(rowData(dat.orig)$crispr[x][c(rP, rN)]))
ha_row = rowAnnotation(df = data.frame(Category = rowData(dat.orig)[x,"crispr"][c(rP,rN)]),
                       col = list(Category = rowcol))

ha_column_crispr = HeatmapAnnotation(df = data.frame(Type = colData(dat.orig)[, c("type")],
                                              Virus = colData(dat.orig)[, c("virus")],
                                              ID = colData(dat.orig)[, c("mcc.id")]),
                              col = list(Type = c(Cell_line="lightgrey", Tumor="black"), 
                                         Virus = c(pos = "#543005", neg = "#DFC27D"),
                                         ID = cols2))

# cluster rows/cols
ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column_crispr,
             row_names_gp = gpar(fontsize = 4),
             col = ecolors)
draw(ht + ha_row)

# don't cluster rows/cols
ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column_crispr,
             row_names_gp = gpar(fontsize = 4),
             col = ecolors,
             cluster_columns = FALSE,
             column_order = seq_len(ncol(dat.orig)),
             cluster_rows = FALSE)
draw(ht + ha_row)
```

Next for vs-transformed counts.

```{r heatmap-crispr-vs, fig.width=9.5, fig.height=14, fig.path='plots/', dev=c('png')}
# subset counts of transformed (and normalized) counts
mat <- assay(vst(dat.orig))[x,][c(rP, rN),]
rownames(mat) <- rowData(dat.orig)$symbol[x][c(rP, rN)]
colnames(mat) <- with(colData(dat.orig), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column_crispr,
             row_names_gp = gpar(fontsize = 3),
             col = ecolors)
draw(ht + ha_row)

# don't cluster rows/cols
ht = Heatmap(mat, name = "VS", 
             top_annotation = ha_column_crispr,
             row_names_gp = gpar(fontsize = 4),
             col = ecolors,
             cluster_columns = FALSE,
             column_order = seq_len(ncol(dat.orig)),
             cluster_rows = FALSE)
draw(ht + ha_row)
```



# MCC Polyomavirus

Here we quantify the presence of Merkel Cell Polyomavirus RNA, as well as visualize where it is located within the MCC genome. In addition, we are interested in identifying specific viral sequences that are highly expressed througout the virus positive samples.

## Quantification of Viral RNA

Here we examine the quantification of the number of reads mapping to viral RNA.
This was carried out using the script in `../../PREPROCESS/RNA/viral/20181029-mapviral.sh`,
which extracts unmapped reads from the RNA-seq bam files and aligns them to the 
reference sequence for the MCC polyomavirus.

```{r, viral}

if(file.exists(viralquant_file)){
 viral <- read.table(viralquant_file) %>%
   mutate(sampleName = gsub("/n/irizarryfs01_backed_up/kkorthauer/MCC/DATA/RNA/", "", V1),
          sampleName = gsub("-", "\\.", gsub(".01.unmapped.virusmap.bam", "", sampleName)),
          dfci.id = gsub("\\.", "_", substr(sampleName, 1, 9)),
          type = ifelse(grepl("T", sampleName), "Tumor", "Cell_line")) %>%
   rename(V2 = "count") %>%
   dplyr::select(-V1, -V3)
   
 colDat$sampleName <- as.character(colDat$sampleName)
 colDat$dfci.id <- as.character(colDat$dfci.id)
 colDat$type <- as.character(colDat$type)
 viral <- left_join(colDat, viral, by = c("sampleName", "dfci.id", "type")) %>%
   mutate(count = ifelse(is.na(count), 0, count))
 
 # normalize by size factors computed earlier  
 sf <- data.frame(sf = sizeFactors(dat))
 sf$sampleName <- rownames(sf) 
 viral <- left_join(viral, sf, by = "sampleName") %>%
   mutate(normcount = count / sf)
 
 p <- ggplot(viral, aes(x = mcc.id, y = count, fill = type)) +
   geom_col(position = "dodge") +
   xlab("MCC ID") +
   ylab("MCC Polyomavirus RNA count (normalized)")+
   ggtitle("Reads mapping to viral RNA")
 
 print(p)
 
 ggsave(file.path("plots", "viralRNAcounts.pdf"), width = 5, height = 3.5)
}
```

## Visualize alignment of MCC RNA

Here we use the `Gviz` package to visualize the alignment of the reads that map to the MCC viral genome. 

```{r track1, fig.height=8, fig.path='plots/', dev=c('png'), eval=FALSE}
coverageTracks <- list.files(path = "/n/irizarryfs01/kkorthauer/MCC/DATA/RNA/", 
                             pattern = ".bedGraph",
                             full.names = TRUE)
covlist <- vector(mode="list", length(coverageTracks ))
covlist <- GRangesList()
track <- vector("list", length(coverageTracks)+1)

nms <- list.files(path = "/n/irizarryfs01/kkorthauer/MCC/DATA/RNA/", 
                  pattern = ".bedGraph")
nms <- sapply(strsplit(nms, "-01_unmapped_"), function(x) x[1])
names(track) <- nms

viral <- unique(viral)
viral$id <- gsub("\\.", "-", viral$sampleName)

id.map <- colData(dat) %>% 
           data.frame() %>%
           dplyr::select(sampleName, mcc.id) %>%
           mutate(id = gsub("\\.", "-", sampleName))
x <- match(names(track), id.map$id)
id.map <- id.map %>%
  mutate(mcc.id.type = paste0(mcc.id, "-", substr(id.map$sampleName, 11,11)))
names(track) <- id.map$mcc.id.type[x]
x <- match(id.map$id, viral$id)
id.map$virus_count <- viral$normcount[x]


options(ucscChromosomeNames=FALSE)
for (i in seq_along(coverageTracks)){
  covlist[[i]] <- import.bedGraph(coverageTracks[i])
  strand(covlist[[i]]) <- "+"
  normfactor <- id.map$virus_count[id.map$mcc.id.type == names(track)[i]]
  score(covlist[[i]]) <- score(covlist[[i]])/normfactor*1e3
  if(length(covlist[[i]]) > 0)
    track[[i]] <- DataTrack(covlist[[i]], name = names(track)[i])
}

track[[i+1]] <- GenomeAxisTrack(name = "MCC")
names(track)[length(track)] <- "MCC"
whichnull <- which(sapply(track, is.null))
track <- track[-whichnull]
covlist <- covlist[-whichnull]

plotTracks(track, type="l" )
```

## Identify highly expressed MCC sequences

Here we use the signal track objects from the previous section to find regions where the MCC expression is above some cutoff for all viral positive samples.

```{r, eval=FALSE}
peaks <- covlist
for (i in seq_along(covlist)){
  peaks[[i]] <- reduce(covlist[[i]][score(covlist[[i]]) > 80])
  if (i == 1)
    int <- peaks[[i]]
  if (i > 1)
    int <- intersect(int, peaks[[i]])
}

# highly expressed sequences in all V+ samples:
int
```

Add these to the coverage track:

```{r track2, fig.height=8, fig.path='plots/', dev=c('png'), eval=FALSE}
mcols(int)$symbol <- c("1", "2")
grtrack <- GeneRegionTrack(int, name = "highly expr")
track <- c(track, grtrack)
track <- track[c(1:(length(track)-2), length(track), length(track)-1)]

plotTracks(track, type="l" )
```

Extract these two sequences from the MCC viral genome:

```{r, eval=FALSE}
require('seqinr')

# Create objects
mygenome <- read.fasta('../../PREPROCESS/RNA/viral/ref/mcc.fa')[[1]]
mcols(int)$seq <- c('seq1','seq2')

myseqs <- lapply(1:length(int), function(x) mygenome[start(int[x,]):end(int[x,])])

# write to a file
write.fasta(myseqs, mcols(int)$seq, 'results/highly_expressed_mcc_rna.fa')

```

We also want to write out these sequences with the two major variants seen in IGV.
There, we see that position 213 changed from C to T, and position 1,205 changed from A to C.

```{r MCCvirustrack, fig.height=8, fig.path='plots/', dev=c('png'), eval=FALSE}
mygenome[213] <- "t"
mygenome[1205] <- "c"

myseqs <- lapply(1:length(int), function(x) mygenome[start(int[x,]):end(int[x,])])

# write to a file
write.fasta(myseqs, mcols(int)$seq, 'results/highly_expressed_mcc_rna_variant.fa')


# add to plot
var <- int
start(var) <- end(var) <- c(213, 1205)
mcols(var)$symbol <- c("C>T", "A>C")

grtrack <- GeneRegionTrack(var, name = "variants")
track <- c(track, grtrack)
track <- track[c(1:(length(track)-2), length(track), length(track)-1)]

plotTracks(track, type="l", transcriptAnnotation = "symbol")
```


# Differential expression

Now we'll investigate differences in expression between cell lines and tumors.

## Difference between tumor and cell line

We'll use DESeq2 to perform standard differential expression between the 
tumor samples and cell lines. We'll exclude the 286 sample as it appears to be
quite an outlier.

```{r}
design(dat) <- ~ W + virus + type
dat <- DESeq(dat)

res <- results(dat)
vsd = vst(dat)

sum(res$padj < 0.01, na.rm = TRUE)

data.frame(res) %>%  
  filter(!is.na(padj)) %>% 
  mutate(pvalue = -log10(pvalue)) %>%
  mutate(DE.01 = padj < 0.01) %>%
  ggplot(aes(log2FoldChange, pvalue, color = DE.01)) +
    geom_point(alpha = 0.4) +
    ylab("-log10(p-value)") + 
    scale_colour_manual(values = c("black", "darkorange"))
ggsave(file.path("plots", "volcano_de01.pdf"), width = 5, height = 4)
```

There are `r sum(res$padj < 0.01, na.rm = TRUE)` DE genes at FDR 0.01. 
`r sum(res$padj < 0.01 & abs(res$log2FoldChange) > 2, na.rm = TRUE)` also have log2 fold change of at least 2. 
Here is a heatmap of those genes, first for log2 normalized counts:

```{r heatmap-DE-l2n-0.01, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
mat <- log2(counts(dat, normalize=TRUE)[res$padj < 0.01 & abs(res$log2FoldChange) > 2 & !is.na(res$padj), ] + 1)
colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column,
             show_row_names = FALSE,
             clustering_distance_rows = "euclidean",
             clustering_distance_columns = "euclidean",
             col = ecolors)
draw(ht)
``` 

Next for vs-transformed counts.

```{r heatmap-DE-vs-0.01, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
mat <- assay(vsd)[res$padj < 0.01 & abs(res$log2FoldChange) > 2 & !is.na(res$padj), ]
colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column,
             show_row_names = FALSE,
             clustering_distance_rows = "euclidean",
             clustering_distance_columns = "euclidean",
             col = ecolors)
draw(ht)
``` 

### GSEA

To understand what groups of genes exist in this set, we'll perform a Gene-set
Enrichment analysis. First well use the biomaRt package to download and 
associate GO annotations for each gene. 

```{r gocat, message=FALSE}
mart <- useMart("ensembl", "hsapiens_gene_ensembl")
goSets <- getBM(c("ensembl_gene_id", "go_id"), mart = mart,
				filters = "ensembl_gene_id", 
				values = unlist(lapply(strsplit(rownames(res), "\\."), 
				                       function(x) x[1])))
desc <- Term(goSets$go_id)
goSets$go_id <- paste0(goSets$go_id, " - ", desc)
goSets <- goSets[!nchar( goSets$go_id ) == 0, ]
goSets <- with(goSets, split(go_id, ensembl_gene_id))
```

Now we'll perform the GSEA with fgsea

```{r}
gsea_file <- file.path("results", "gsea.rds")
goSets <- split(rep(names(goSets), lengths(goSets)), unlist(goSets))
stats <- res$stat
names(stats) <- unlist(lapply(strsplit(rownames(res), "\\."), 
				                       function(x) x[1]))
stats <- stats[!is.na(stats)]
overwrite <- TRUE

if (!file.exists(gsea_file) || overwrite) {
  goRes <- fgsea(goSets, 
                  stats, 
                  nperm=10000, 
                  maxSize=500,
                  minSize=5)  
  saveRDS(goRes, file = gsea_file)
} else {
  goRes <- readRDS(gsea_file)
}

head(goRes[order(pval), ])
sum(goRes[, padj < 0.01])
```

Here we visualize the top pathways:

```{r gsea, fig.width=8, fig.height=9, fig.path='plots/', dev=c('png'), warning = FALSE, message = FALSE}
topPathwaysUp <- goRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- goRes[ES < 0][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(goSets[topPathways], stats, goRes, 
              gseaParam = 0.5,  
              colwidths = c(9, 4.5, 1.2, 1.7, 1.7))
```

The top categories that DE genes between Tumor and cell line are overrepresented
are highly related to the immune response. 

Since many of these pathways are related and overlapping, we also try collapsing
into independent pathways:

```{r gsea-collapsed, fig.width=8, fig.height=9, fig.path='plots/', warning = FALSE, message = FALSE, dev=c('png')}
collapsedPathways <- collapsePathways(goRes[order(pval)][padj < 0.01], 
                                      goSets, stats)
mainPathways <- goRes[pathway %in% collapsedPathways$mainPathways][
                         order(-NES), pathway][1:20]
plotGseaTable(goSets[mainPathways], stats, goRes, 
              gseaParam = 0.5,
              colwidths = c(9, 4.5, 1.2, 1.7, 1.7))
```

### Overlap with immune Genes of Interest

Here we examine the fold changes and significance of differential expression between
the tumor and cell lines for the genes of interest.

```{r}
df <- data.frame(res) %>%  
  mutate(description = rowData(dat)$description,
         symbol = rowData(dat)$symbol) %>%
  filter(!is.na(padj)) %>% 
  mutate(pvalue = -log10(pvalue)) %>%
  mutate(DE.01 = padj < 0.01)
  
knitr::kable((df %>% filter(!is.na(description)) %>% arrange(-pvalue))[,-c(1,3,4,5)])
```

`r sum(df$DE.01 & !is.na(df$description))` out of 
`r sum(!is.na(df$description))` genes of interest (
`r signif(100*sum(df$DE.01 & !is.na(df$description))/sum(!is.na(df$description)),3)`%) 
are DE at the 0.01 FDR level. This is compared to 
`r signif(100*sum(df$DE.01)/nrow(df),3)`% genes overall
(`r sum(df$DE.01)` out of 
`r nrow(df)`). 

In particular, note that this quantity broken down by category is:

* `r sum(df$DE.01 & df$description == "MCC Marker" & !is.na(df$description))` out of 
`r sum(df$description == "MCC Marker" & !is.na(df$description))` MCC Marker genes 
(`r signif(100*sum(df$DE.01 & !is.na(df$description) & df$description == "MCC Marker")/sum(df$description == "MCC Marker" & !is.na(df$description)),3)`%)

* `r sum(df$DE.01 & df$description == "MAPK Pathway" & !is.na(df$description))` out of 
`r sum(df$description == "MAPK Pathway" & !is.na(df$description))` MAPK Pathway genes 
(`r signif(100*sum(df$DE.01 & !is.na(df$description) & df$description == "MAPK Pathway")/sum(df$description == "MAPK Pathway" & !is.na(df$description)),3)`%)

* `r sum(df$DE.01 & df$description == "Class I" & !is.na(df$description))` out of 
`r sum(df$description == "Class I" & !is.na(df$description))` Class I genes 
(`r signif(100*sum(df$DE.01 & !is.na(df$description) & df$description == "Class I")/sum(df$description == "Class I" & !is.na(df$description)),3)`%)

* `r sum(df$DE.01 & df$description == "Class II" & !is.na(df$description))` out of 
`r sum(df$description == "Class II" & !is.na(df$description))` Class II genes 
(`r signif(100*sum(df$DE.01 & !is.na(df$description) & df$description == "Class II")/sum(df$description == "Class II" & !is.na(df$description)),3)`%)

We'll visualize the genes of interest on the volcano plot.

```{r volcanogoi, fig.path='plots/', dev=c('png')}
p <- ggplot(data = df, aes(log2FoldChange, pvalue, color = DE.01)) +
    geom_point(alpha = 0.4) +
    ylab("-log10(p-value)") + 
    scale_colour_manual(values = c("black", "darkorange")) +
    geom_point(data = df %>% filter(!is.na(description)), color = "red") +
    geom_label_repel(data = df %>% filter(!is.na(description)) %>% filter(DE.01), 
               aes(label = symbol), color = "black", size = 3)
p
ggsave(file.path("plots", "volcano_de01_goi.pdf"), plot = p, width = 7, height = 5)
```

## Difference between viral positive and negative

We'll use DESeq2 to perform standard differential expression between the 
viral positive and viral negative samples. We'll include cell line / tumor
origin as a covariate.

```{r volcanoviral, fig.path='plots/', dev=c('png')}
design(dat) <- ~ W + type + virus
dat <- DESeq(dat)

res <- results(dat)
vsd = vst(dat)

sum(res$padj < 0.01, na.rm = TRUE)

data.frame(res) %>%  
  filter(!is.na(padj)) %>% 
  mutate(pvalue = -log10(pvalue)) %>%
  mutate(DE.01 = padj < 0.01) %>%
  ggplot(aes(log2FoldChange, pvalue, color = DE.01)) +
    geom_point(alpha = 0.4) +
    ylab("-log10(p-value)") + 
    scale_colour_manual(values = c("black", "darkorange"))
ggsave(file.path("plots", "volcano_de01_viral.pdf"), width = 5, height = 4)
```

There are `r sum(res$padj < 0.01, na.rm = TRUE)` DE genes at FDR 0.01. 
`r sum(res$padj < 0.01 & abs(res$log2FoldChange) > 2, na.rm = TRUE)` also have log2 fold change of at least 2. 
Here is a 
heatmap of those genes, first for log2 normalized counts:

```{r heatmap-DE-l2n-0.01-viral, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
mat <- log2(counts(dat, normalize=TRUE)[res$padj < 0.01 & !is.na(res$padj) & abs(res$log2FoldChange) > 2, ] + 1)
colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "log2\nNormed\ncounts", 
             top_annotation = ha_column,
             show_row_names = FALSE,
             clustering_distance_rows = "euclidean",
             clustering_distance_columns = "euclidean",
             col = ecolors)
draw(ht)
``` 

Next for vs-transformed counts.

```{r heatmap-DE-vs-0.01-viral, fig.width=8.5, fig.height=7, fig.path='plots/', dev=c('png')}
mat <- assay(vsd)[res$padj < 0.01 & !is.na(res$padj) & abs(res$log2FoldChange) > 2, ]
colnames(mat) <- with(colData(dat), paste0(mcc.id, " ", gsub("_", " ", type))) 

ht = Heatmap(mat, name = "VS counts", 
             top_annotation = ha_column,
             show_row_names = FALSE,
             clustering_distance_rows = "euclidean",
             clustering_distance_columns = "euclidean",
             col = ecolors)
draw(ht)
``` 

### GSEA

To understand what groups of genes exist in this set, we'll perform a Gene-set
Enrichment analysis. First well use the biomaRt package to download and 
associate GO annotations for each gene. 

```{r gocat-viral, message=FALSE}
mart <- useMart("ensembl", "hsapiens_gene_ensembl")
goSets <- getBM(c("ensembl_gene_id", "go_id"), mart = mart,
				filters = "ensembl_gene_id", 
				values = unlist(lapply(strsplit(rownames(res), "\\."), 
				                       function(x) x[1])))
desc <- Term(goSets$go_id)
goSets$go_id <- paste0(goSets$go_id, " - ", desc)
goSets <- goSets[!nchar( goSets$go_id ) == 0, ]
goSets <- with(goSets, split(go_id, ensembl_gene_id))
```

Now we'll perform the GSEA with fgsea

```{r}
gsea_file <- file.path("results", "gsea-viral.rds")
goSets <- split(rep(names(goSets), lengths(goSets)), unlist(goSets))
stats <- res$stat
names(stats) <- unlist(lapply(strsplit(rownames(res), "\\."), 
				                       function(x) x[1]))
stats <- stats[!is.na(stats)]

if (!file.exists(gsea_file) || overwrite) {
  goRes <- fgsea(goSets, 
                  stats, 
                  nperm=10000, 
                  maxSize=500,
                  minSize=5)  
  saveRDS(goRes, file = gsea_file)
} else {
  goRes <- readRDS(gsea_file)
}

head(goRes[order(pval), ])
sum(goRes[, padj < 0.05])
```

Here we visualize the top pathways:

```{r gsea-viral, fig.width=8, fig.height=9, fig.path='plots/', dev=c('png'), warning = FALSE, message = FALSE}
topPathwaysUp <- goRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- goRes[ES < 0][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(goSets[topPathways], stats, goRes, 
              gseaParam = 0.5,  
              colwidths = c(9, 4.5, 1.2, 1.7, 1.7))
```

The top categories that DE genes between viral positive and viral negative are overrepresented
are highly related to the immune response. 

Since many of these pathways are related and overlapping, we also try collapsing
into independent pathways:

```{r gsea-collapsed-viral, fig.width=8, fig.height=9, fig.path='plots/', warning = FALSE, message = FALSE, dev=c('png')}
collapsedPathways <- collapsePathways(goRes[order(pval)][padj < 0.05], 
                                      goSets, stats)
mainPathways <- goRes[pathway %in% collapsedPathways$mainPathways][
                         order(-NES), pathway][1:20]
plotGseaTable(goSets[mainPathways], stats, goRes, 
              gseaParam = 0.5,
              colwidths = c(9, 4.5, 1.2, 1.7, 1.7))
```

### Overlap with immune Genes of Interest

Here we examine the fold changes and significance of differential expression between
the tumor and cell lines for the genes of interest.

```{r}
df <- data.frame(res) %>%  
  mutate(description = rowData(dat)$description,
         symbol = rowData(dat)$symbol) %>%
  filter(!is.na(padj)) %>% 
  mutate(pvalue = -log10(pvalue)) %>%
  mutate(DE.01 = padj < 0.01)
  
knitr::kable((df %>% filter(!is.na(description)) %>% arrange(-pvalue))[,-c(1,3,4,5)])

```

`r sum(df$DE.01 & !is.na(df$description))` out of 
`r sum(!is.na(df$description))` genes of interest (
`r signif(100*sum(df$DE.01 & !is.na(df$description))/sum(!is.na(df$description)),3)`%) 
are DE at the 0.01 FDR level. This is compared to 
`r signif(100*sum(df$DE.01)/nrow(df),3)`% genes overall
(`r sum(df$DE.01)` out of 
`r nrow(df)`). 

In particular, note that this quantity broken down by category is:

* `r sum(df$DE.01 & df$description == "MCC Marker" & !is.na(df$description))` out of 
`r sum(df$description == "MCC Marker" & !is.na(df$description))` MCC Marker genes 
(`r signif(100*sum(df$DE.01 & !is.na(df$description) & df$description == "MCC Marker")/sum(df$description == "MCC Marker" & !is.na(df$description)),3)`%)

* `r sum(df$DE.01 & df$description == "MAPK Pathway" & !is.na(df$description))` out of 
`r sum(df$description == "MAPK Pathway" & !is.na(df$description))` MAPK Pathway genes 
(`r signif(100*sum(df$DE.01 & !is.na(df$description) & df$description == "MAPK Pathway")/sum(df$description == "MAPK Pathway" & !is.na(df$description)),3)`%)

* `r sum(df$DE.01 & df$description == "Class I" & !is.na(df$description))` out of 
`r sum(df$description == "Class I" & !is.na(df$description))` Class I genes 
(`r signif(100*sum(df$DE.01 & !is.na(df$description) & df$description == "Class I")/sum(df$description == "Class I" & !is.na(df$description)),3)`%)

* `r sum(df$DE.01 & df$description == "Class II" & !is.na(df$description))` out of 
`r sum(df$description == "Class II" & !is.na(df$description))` Class II genes 
(`r signif(100*sum(df$DE.01 & !is.na(df$description) & df$description == "Class II")/sum(df$description == "Class II" & !is.na(df$description)),3)`%)

We'll visualize the genes of interest on the volcano plot.

```{r volcanogoiviral, fig.path='plots/', dev=c('png')}
p <- ggplot(data = df, aes(log2FoldChange, pvalue, color = DE.01)) +
    geom_point(alpha = 0.4) +
    ylab("-log10(p-value)") + 
    scale_colour_manual(values = c("black", "darkorange")) +
    geom_point(data = df %>% filter(!is.na(description)), color = "red") +
    geom_label_repel(data = df %>% filter(!is.na(description)) %>% filter(DE.01), 
               aes(label = symbol), color = "black", size = 3)
p
ggsave(file.path("plots", "volcano_de01_goi_viral.pdf"), plot = p, width = 7, height = 5)
```
# Session Information

```{r}
sessionInfo()
```
